---
title: "ODSC 2016 ggplot2 workshop"
author: "Jared Lander"
date: "May 21, 2016"
output: html_document
---

“Workshop:ggplot2: from scratch to compelling graphs”


the simple way of plotting in R
`plot(price ~ carat, data=diamonds)`
`hist(diamonds$price)`
`boxplot(diamonds$price)`

using `ggplot2` with R studio's built in `diamonds` dataset
```{r echo=FALSE} 
library(ggplot2)
library(ggplot2)
library(ggthemes)

data(diamonds)
head(diamonds)

ggplot(diamonds, aes(x=carat, y=price)) + geom_point()
```


### set a plot's aesthetics with `aes()`
map the diamond's `cut` value to the color aesthetic
`aes(color=cut)`
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point(aes(color=cut))
```

or hard code the color variable
`color='blue'`
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point(color='blue')
```

### histograms
`geom_histagram` automatically calcualtes heights. No Y value necessary
change resolution with the aesthetics `binwidth` argument
```{r}
ggplot(diamonds) + geom_histogram(aes(x=price), binwidth = 100)
```

color property sets border on 2d objects, fill sets fill color
`fill='red', color='blue'
```{r}
ggplot(diamonds) + geom_histogram(aes(x=price), fill='red', color='blue')
```


###compare the aesthetics of the following 2 examples:
when you specify aes() inside ggplot function, it applies to all layers.
in this case we have color assigned to the diamond's `cut` property in ggplot's `aes()` function like so... `ggplot(diamonds, aes(x=carat, y=price, color=cut))`
```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + 
  geom_point(aes(color=cut)) + 
  geom_smooth()
```

while in this case, color is assigned to `cut`, but only in the `geom_point()` function's aesthetic. 
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + 
  geom_point(aes(color=cut)) +
  geom_smooth()
```

### transparency
transparency is achieved here by setting `alpha=1/3` in `geom_point()`
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + 
  geom_point(aes(color=cut), shape=1, size=2, alpha=1/3) +
  geom_smooth()

```


### multiples aka facets
assign base function to a variable! 
```{r}
g <- ggplot(diamonds, aes(x=carat, y=price, color=cut))
g + geom_point() + facet_wrap( ~ cut)
```

1d row or col of facets
```{r}
g + geom_point() + facet_wrap( ~ cut, nrow=1, scales='free') #also try nrow=1
```

2d grid of facets
```{r}
g + geom_point() + facet_grid(color ~ clarity)
```

###scales
scales='free' argument allows the scale of the axis of individual facets to vary
scales='free_y' allows free y axis and fixed x axis

using free scales is totally misleading, 
at a glance suggests all cuts of diamonds are the same price

research forumla notation (use of ~ )....


***compare the 2 following examples***
  
```{r}
ggplot(diamonds, aes(x=price)) + geom_histogram() + 
  facet_wrap(~cut)
```


```{r}
ggplot(diamonds, aes(x=price)) + geom_histogram() + 
  facet_wrap(~cut, scales = 'free')
```



###boxplots...
```{r}
ggplot(diamonds, aes(x=1, y=price)) + geom_boxplot()
ggplot(diamonds, aes(x=cut, y=price)) + geom_boxplot()
```
..vs 
###violins (more informative!)
```{r}
ggplot(diamonds, aes(x=cut, y=price)) + geom_violin()
```

***more on violin plots...***

when we add geom_point, the x axis is discrete, so not very helpful
```{r}
ggplot(diamonds, aes(x=cut, y=price)) + geom_violin() +
  geom_point()
```

usingjittering will make the points more useful to show their density.
also, the order matters! put the violin on top layer (last) to make it more useful

```{r}
ggplot(diamonds, aes(x=cut, y=price)) +
  geom_jitter(alpha=1/4) +
  geom_violin(alpha=1/2, draw_quantiles = .5) 
```

```{r}
g1 <- ggplot(diamonds, aes(x=carat, y=price)) +
  geom_point(aes(color=cut))
```

# style your plots with gg_themes
```{r}
g1 + theme_economist()
g1 + theme_fivethirtyeight()
g1 + theme_wsj() + scale_color_wsj()
g1 + theme_bw()
```

###labels
```{r}
g2 <-  ggplot(diamonds, aes(x=carat, y=price)) +
  geom_point()

g2 + labs(x='Carat', y='Price ($)', title='Price by Carat')
```

alternatively
```{r}
g2 + xlab('Carat') + ylab('Price ($)') + ggtitle('Price by Carat')
```

###transformations
add dollar sign to labels
```{r}
g2 + labs(x='Carat', y='Price', title='Price by Carat') + 
  scale_y_continuous(label=scales::dollar)
```


comma delimit numbers
***note the :: syntax allows you to use function from package you haven't loaded***
```{r}
g2 + labs(x='Carat', y='Price ($)', title='Price by Carat') + 
  scale_y_continuous(label=scales::comma)  

```


brewer is a color scale for color blind (i think)
(see other scale_color_ functions)
```{r}
g1 + scale_color_brewer()
```

move the legend to the bottom
```{r}
g1 + theme(legend.position='bottom')
```


###zooming
xlim zooms by removing data.
this is problematic because the smoothing curve no longer has certain values, thus the output is distorted

```{r}
g3  <- g2 + geom_smooth()
g3 + xlim(c(0, 3))
```

so how can we just zoom?
```{r}
g3 + coord_cartesian(xlim=c(1, 3))
```

rotate 90 degrees
```{r}
g3 + coord_flip()
```

polar coordinates
```{r}
g3 + coord_polar()
```


an ugly heatmap
```{r}
library(scales)
library(tidyr)
library(reshape2)
library(dplyr)

# economic indicators dataset
head(economics)

#correlation matrix between these specific variables
econCor <- cor(economics[, c(2, 4:6)]) #this is in "y format" 


# todo learn about piping in R %<%  (cmd+shift+ m)
econMelt <- melt(econCor, varnames=c('x', 'y'), value.name='Correlation')

head(econMelt)

econMelt <- econMelt %>%  arrange(Correlation)

head(econMelt)

# set heatmap to h
h <- ggplot(econMelt, aes(x=x, y=y)) + geom_tile(aes(fill=Correlation))

#now style the heatmap

h + scale_fill_gradient2(low=muted('red'), mid='white', high='steelblue',
  guide=guide_colorbar(ticks=FALSE, barheight=10), limits=c(-1, 1)) + 
  theme_minimal() + labs(x=NULL, y=NULL) 

```

#todo learn about dplyer


